---
title: "Midterm"
author: Adaeze Obinelo
format: html
editor: visual
---

## Research Question: Am I really paying more to fly home for thanksgiving than anyone ever has before?

Background: Every year it seems like buying my ticket home for the holidays hurts a little bit more than it did the year before. Am I being dramatic, or is the cost of airfare actually going up compared to years prior?

[Precise research question:]{.underline} **Is flying round trip from LAX to Logan Airport(BOS) more expensive on average than it has been in previous decades, when adjusting for 2023 inflation?**

## Step 1: Get my data, set up my library

[**Data Source:**]{.underline}

I took data available from the second financial quarter of 4 years, 1993, 2003, 2013, and 2023; I felt that a ten year spread would be good enough to get a general sense of fare trends.

The names of some of the variables were overall very clunky, so I tried to fix that here.

Additionally had to change the coding to start after the first row, as below

```{r}
library(dplyr)
library(data.table)
library(tidyverse)
library(ggplot2)
library(readxl)

set1 <- read_excel('1993f.xls')
set2 <- read_excel('2003f.xls')
set3 <-read_excel('2013f.xls')
set4 <- read_excel('2023f.xls')

names(set1) <- set1[1,] # the data loaded with "header" included as data, so fixing that
set1 <- set1[-1,]

names(set2) <- set2[1,] 
set2 <- set2[-1,]

names(set3) <- set3[1,] 
set3 <- set3[-1,]

names(set4) <- set4[1,] 
set4 <- set4[-1,]

```

```{r}
names(set1)[names(set1) == "Average Fare ($)"] <- "Avg_1993"
colnames(set1)[7]<- "Adjusted_Avg_Fare_Q2_1993"
colnames(set1)[2]<- "Airport"
colnames(set1)[1]<- "PassRank_1993"
colnames(set1)[8]<- "_1993v2022_Passengers"

names(set2)[names(set2) == "Average Fare ($)"] <- "Avg_2003"
colnames(set2)[7]<- "Adjusted_Avg_Fare_Q2_2003"
colnames(set2)[2]<- "Airport"
colnames(set2)[1]<- "Pass_Rank_2003"
colnames(set2)[8]<- "_2003v2022_Passengers"

names(set3)[names(set3) == "Average Fare ($)"] <- "Avg_2013"
colnames(set3)[7]<- "Adjusted_Avg_Fare_Q2_2013"
colnames(set3)[2]<- "Airport"
colnames(set3)[1]<- "Pass_Rank_2013"
colnames(set3)[8]<- "_2013v2022_Passengers"

names(set4)[names(set4) == "Average Fare ($)"] <- "Avg_2023"
colnames(set4)[7]<- "Adjusted_Avg_Fare_Q2_2023"
colnames(set4)[2]<- "Airport"
colnames(set4)[1]<- "Pass_Rank_2023"
colnames(set4)[8]<- "_2023v2022_Passengers"

```

Combined my data into one dataset below. I realized the data was coded as character variables, so had to change my fare variables to dbl format as below.

```{r}
trx1 <- merge(set1, set2, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))
dim(trx1)

trx2 <- merge(set3, set4, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))
dim(trx2)

trx <- merge(trx1, trx2, by=c('Airport', 'Airport Name', 'City Name', 'State Name'))

trx$PassRank_1993 = as.numeric(as.character(trx$PassRank_1993))
trx$Adjusted_Avg_Fare_Q2_1993 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_1993))
trx$Pass_Rank_2003 = as.numeric(as.character(trx$Pass_Rank_2003))
trx$Adjusted_Avg_Fare_Q2_2003 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2003))

trx$Pass_Rank_2013 = as.numeric(as.character(trx$Pass_Rank_2013))
trx$Adjusted_Avg_Fare_Q2_2013 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2013))
trx$Pass_Rank_2023 = as.numeric(as.character(trx$Pass_Rank_2023))
trx$Adjusted_Avg_Fare_Q2_2023 = as.numeric(as.character(trx$Adjusted_Avg_Fare_Q2_2023))

trx$Avg_1993 = as.numeric(as.character(trx$Avg_1993))
trx$Avg_2003 = as.numeric(as.character(trx$Avg_2003))
trx$Avg_2013 = as.numeric(as.character(trx$Avg_2013))
trx$Avg_2023 = as.numeric(as.character(trx$Avg_2023))

library(purrr)
trx<-modify_if(trx, ~is.numeric(.), ~round(., 3))


trx <- na.omit(trx)
```

The bottom of our dataset has statistics that I don't need --\> will take care of them below

## Added coordinates just in case

I like having coordinate data just in case I'll need it (maps are fun), so I found a data set with lat and longitude coordinates for US airports

-   pulled the info I needed from this dataset, then merged by airport abbreviation

```{r}
airports <- read_csv('us-airports.csv')

gps <- airports %>% 
  select(local_code, latitude_deg, longitude_deg)

names(gps) <- gps[1,] 
gps <- gps[-1,]
colnames(gps)[1]<- "Airport"
colnames(gps)[2]<- "lat"
colnames(gps)[3]<- "lon"


trx <- merge(trx, gps, by=c('Airport'))

trx <- trx %>% drop_na(`Airport Name`) #drop rows missing aiport name (from og trx)'

trx$lat = as.numeric(as.character(trx$lat))
trx$lon = as.numeric(as.character(trx$lon))

```

Dataset is smaller after dropping NA's only bc the bottom of the dataset has BoTX statistics ( I dont need them).

## Step 2: Check Data:

I felt like the best way to check my data would be to compare the inflation adjusted fare variable for each year (Adjusted_Avg_Fare_Q2) provided originally with my dataset with an actual dollar to dollar conversion between that year and 2023.

Conversions:

1993: 1-\>2.13

2003: 1-\>1.67

2013: 1-\>1.32

-   I coded a new variable "check.adj(xx)" for each year and compared the value

```{r}
check <- trx %>% 
  select(Avg_1993, Avg_2003, Avg_2013, Avg_2023) %>% 
  mutate(check.adj.93 = (Avg_1993) *2.13,
                  check.adj.03 = (Avg_2003) * 1.67,
                  check.adj.13 = (Avg_2013) * 1.32,
            check.adj.23 = (Avg_2023)* 1)

trx <- merge(trx, check, by=c('Avg_1993', 'Avg_2003', 'Avg_2013', 'Avg_2023'))

```

```{r}

trx %>% 
  select(Adjusted_Avg_Fare_Q2_1993,check.adj.93, Adjusted_Avg_Fare_Q2_2003, check.adj.03, Adjusted_Avg_Fare_Q2_2013, check.adj.13, Adjusted_Avg_Fare_Q2_2023, check.adj.23)%>% 
  slice(1:5)
   
```

The value of my check variable and the actual inflation adjusted fare isn't a perfect match, but its very close so I would consider my data valid.

## Step 3: Bird's eye view --\> Comparing Airport Fare to Passenger Rank

The dataset includes a variable Passenger Rank, that is compiled by comparing the overall passenger traffic at that airport for that year to the overall amount of total passengers at all airports in 2022. This variable allows us to rank the passenger load of that airport for that year to the rest in that year.

Are there any broad trends in the data provided? Logan and LAX are high traffic airports- can we predict any associations between their passenger rank and the average fare?

```{r}
trx %>% 
  filter(Airport== "LAX"| Airport== "BOS") %>% 
  select(Airport, Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993) 
  
```

Both LAX and BOS consistently rank among the top 10 for passenger rank for all 4 years.

What can be said about passenger rank and average fare?

```{r}
names(trx)[names(trx) == "State Name"] <- "State"
west <- trx %>%
  filter(State=="CA"|State=="AK"|State=="WA"|State=="ID"|State=="OR"|State=="NV"|State=="UT")

west %>%
  ggplot()+
  geom_point(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013" )) +
  geom_smooth(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013"), method = "lm")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Trend of Inflation Adjusted Fare vs Passenger Rank of Western US Airports in 2003")+
  geom_point(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993")) +
  geom_smooth(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023"), method = "lm")+
  xlab(label = "Passenger Rank") +
  ylab(label = "Inflation Adjusted Average Fare")+
  scale_color_manual(name="legend",values = c("blue","yellow","pink","red"))

  


```

Looking at the trend for all years in my data set, it looks like as the Passenger Rank decreases, broadly speaking so does the cost of the average inflation adjusted flight there in the West.

It also looks like compared to previous years, 2023 fares range lower than the fares from the years prior, excluding a few outliers in the lower passenger rank region of the graph.

```{r}

east <- trx %>%
  filter(State=="NY"|State=="PA"|State=="MD"|State=="DE"|State=="NJ"|State=="CT"|State=="RI"|State=="VT"|State=="MA"|State=="NH"|State=="ME")

east %>%
  ggplot()+
  geom_point(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2003 , y = Adjusted_Avg_Fare_Q2_2003, color = "2003"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013" )) +
  geom_smooth(mapping = aes(x = Pass_Rank_2013 , y = Adjusted_Avg_Fare_Q2_2013, color = "2013"), method = "lm")+
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "Trend of Inflation Adjusted Fare vs Passenger Rank of Western US Airports in 2003")+
  geom_point(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993")) +
  geom_smooth(mapping = aes(x = PassRank_1993 , y = Adjusted_Avg_Fare_Q2_1993, color = "1993"), method = "lm")+
  geom_point(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023")) +
  geom_smooth(mapping = aes(x = Pass_Rank_2023 , y = Adjusted_Avg_Fare_Q2_2023, color = "2023"), method = "lm")+
  xlab(label = "Passenger Rank") +
  ylab(label = "Inflation Adjusted Average Fare")+
  scale_color_manual(name="legend",values = c("blue","yellow","pink","red"))
```

However, it does not look like this trend of is as strong in the east. Again, 2023 fares range lower than in previous decades when comparing inflation adjusted averages. From these figures, its starting to look like I might actually be paying less to fly out of LAX and BOS, two high traffic airports, than I would be in previous years.

How about if I compare Unadjusted average Fares across the four years of data?

```{r}
LAX.BOS<-trx %>% 
  filter(Airport== "LAX"|Airport== "BOS") %>% 
  select(Airport, Adjusted_Avg_Fare_Q2_1993,Adjusted_Avg_Fare_Q2_2003,Adjusted_Avg_Fare_Q2_2013,Adjusted_Avg_Fare_Q2_2023, Avg_1993, Avg_2003, Avg_2013, Avg_2023,Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993)

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Avg_1993, x = Airport), stat= "identity", fill = "blue")+
  xlab(label = "Airport") +
  ylab(label = "Unadjusted Average Fare 1993")

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Avg_2003, x = Airport), stat= "identity", fill = "yellow")+
  xlab(label = "Airport") +
  ylab(label = "Unadjusted Average Fare 2003")

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Avg_2013, x = Airport), stat= "identity", fill = "pink")+
  xlab(label = "Airport") +
  ylab(label = "Unadjusted Average Fare 2013")

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Avg_2023, x = Airport), stat= "identity", fill = "red")+
  xlab(label = "Airport") +
  ylab(label = "Unadjusted Average Fare 2023")

```

Looks as if the cost of the average ticket out of Logan has remained around the same across the 4 years of data that I collected, with a slight downtrend in 2013, which then up trended again into 2023. The average ticket out of LAX on the other-hand seems to have up trended steadily since 1993, starting at under \$350 to a little over \$400 by 2023.

## Step 4: Answering My Question

How, lets look at adjusted fare across the 4 years of data, but just for LAX and BOS

```{r}

LAX.BOS<-trx %>% 
  filter(Airport== "LAX"|Airport== "BOS") %>% 
  select(Airport, Adjusted_Avg_Fare_Q2_1993,Adjusted_Avg_Fare_Q2_2003,Adjusted_Avg_Fare_Q2_2013,Adjusted_Avg_Fare_Q2_2023, Avg_1993, Avg_2003, Avg_2013, Avg_2023,Pass_Rank_2023, Pass_Rank_2013, Pass_Rank_2003, PassRank_1993)

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Adjusted_Avg_Fare_Q2_1993, x = Airport), stat= "identity", fill = "blue")+
  xlab(label = "Airport") +
  ylab(label = "Inflation Adjusted Average Fare 1993")

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Adjusted_Avg_Fare_Q2_2003, x = Airport), stat= "identity", fill = "yellow")+
  xlab(label = "Airport") +
  ylab(label = "Inflation Adjusted Average Fare 2003")

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Adjusted_Avg_Fare_Q2_2013, x = Airport), stat= "identity", fill = "pink")+
  xlab(label = "Airport") +
  ylab(label = "Inflation Adjusted Average Fare 2013")

LAX.BOS %>% 
  ggplot()+
  geom_bar(aes(y=Adjusted_Avg_Fare_Q2_2023, x = Airport), stat= "identity", fill = "red")+
  xlab(label = "Airport") +
  ylab(label = "Inflation Adjusted Average Fare 2023")
  
  

```

Boston: Looks like I'll actually be paying *less* than I would have in prior years, particularly in 1993. Almost \$800 to fly in inflation adjusted rate, compared to \$400 today.

LAX: Not as bad as Logan, which was about \$700 to fly out of in 1993, compared to about \$450 today.

*How much less compared to 1993?*

```{r}

LAX.BOS <- LAX.BOS %>% 
  mutate(
    comp_fare = Adjusted_Avg_Fare_Q2_1993 - Adjusted_Avg_Fare_Q2_2023)

LAX.BOS %>% 
  select(Airport, comp_fare)

278.577	+ 369.449	
```

It looks like on average, I would be paying \$648 more for a round trip in 1993 when comparing inflation adjusted rates. So it looks like, at least comparing the years that I did, I actually have it better off with what I pay now compared to what I would be paying in 1993.

This is however limited scope--\> I only looked at one quarter and across 4 interrupted time points.

## Conclusion- Am I spending more money?

It looks like if we're just looking at the dollar amount, yes I spend slightly more in 2023 than what people spent for the average round trip from BOS to LAX in previous years, primarily thanks to the rising cost of a ticket out of LAX. However, in inflation adjusted dollars (which is more relevant, even though it doesn't feel any better) I am actually spending less. [So in conclusion, I am **not** spending more in relative 2023 dollars than anyone else has every spent before for a round trip between LAX and Boston.]{.underline}

## Extra; What's the most expensive place to fly out of in 2023? What is the least expensive?

```{r}
library(leaflet)
library(devtools)
library(leaflet)
library(magrittr)

trx<-trx %>% 
  arrange(desc(Avg_2023))

max <- trx %>%
  select(Avg_2023, lat, lon, Airport) %>%
  filter(row_number() == 1)

min <- trx %>%
  select(Avg_2023, lat, lon, Airport) %>%
  filter(row_number() == 369)

leaflet() %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(
    data = min,
    lat = ~lat, lng = ~lon, popup = "Min Avg Fare.",
    opacity = 1, fillOpacity = 1, radius = 400, color = "blue"
  ) %>%
  addCircles(
    data = max,
    lat = ~lat, lng = ~lon, popup = "Max Avg Fare.",
    opacity=1, fillOpacity=1, radius = 400, color = "red")
```

```{r}

max

min
```

The most expensive Airport to Fly out of in 2023 is Central Business District in Alaska.

The least expensive to fly out of is Santa Maria Public Airport in Santa Barbara.
